---
title: "Blog Post 1: Analyzing Transactional Data Using Assocation Rules"
author: "Reed Gabriel"
date: "December 6, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The data used in this analysis comes from a chain of bakeries. The 'purchases.csv' file contains the receipts of 75,000 transactions from these bakeries, with each row representing a receipt and each column representing whether or not a product was purchased. A '1' in the column indicates that the product was purchased for that transaction and a '0' in the column indicates that the product was not purchased for that transaction. The other file, 'items.csv', contains a list of the 50 items that can be purchased at the bakeries. The order of the rows for the items in this table corresponds to the order of the columns in the 'purchases.csv' table. So for example, row 1 (the first product) in the 'items.csv' is Chocolate Cake, and the 2nd column in the 'purchases.csv' represents whether or not each purchase contained a Chocolate Cake or not.

```{r}
# Set working directory. Make sure the 2 files are placed in this location on your local system.
setwd("~/BZAN 552 - Multivariate-Data Mining II/FinalPortfolio/Post 1 - Bakery")

# Now we need to first load in the data.
purchases <- read.table("purchases.csv", head = FALSE, sep = ",", stringsAsFactors = FALSE)
items <- read.table("items.csv", head = FALSE, sep = ",", stringsAsFactors = FALSE)

# To better understand each table before continuing, we can take a look at the dimensions and first few rows.
dim(purchases)
head(purchases[1:5,])
# The purchases dataset contains 75,000 rows (transactions) and 51 columns, with the 1st column being the row numbers, and the binary columns following that from column 2 to column 51.

dim(items)
head(items[1:5,])
# The items dataset contains 50 rows (items) and 2 columns, with the 1st column being the row numbers, and the 2nd column showing the name of the item.
```

Now, we need to combine these tables and convert it into transactional form. Although the 'purchases.csv' does list each row as a transaction, each column is a binary variable representing the presence or absence of an item in the purchase. Transactional form gives a list of the names of the items in each transaction separated by a comma. This form is not only simpler to read and understand, but also works well with assocation rule mining techniques.

```{r}
# Removing the first column in the purchases table (contains the row numbers)
purchases$V1 <- NULL
# Naming the columns as the corresponding item from the items table
colnames(purchases) <- items[,2]
# Converting to matrix
purchases <- as.matrix(purchases)
# Converting to transactional form
#install.packages("arules")
library(arules)
trans_data <- as(purchases, "transactions")

# Let's take a look at the first 5 rows of our data in transactional form to make sure it looks correct.
inspect(head(trans_data[1:5]))

# Before applying a Market Basket Analysis, we can look at an item frequency plot to give an idea of which items are being purchased the most frequently for the time period in which this data was collected. I have chosen to look at the top 15 items.
itemFrequencyPlot(trans_data, topN = 15, type = "absolute", horiz = TRUE, col = "skyblue", main  = "Item Frequency Plot", xlab = "Item Frequency (Counts)")
# We see that the top item is the Coffee Eclair, which was found in over 8,000 purchases. A couple other popular items that show up in a significant number of purchases are Hot Coffee and the Tuile Cookie.
```

Before starting our Market Basket Analysis, I would like to take a moment to define three terms that are commonly used when performing association rule mining. These terms are support (sup), confidence (conf), and lift.

**Support:** The support is the fraction of transactions to which the rule may applicable, i.e., the fraction of transactions that have all items mentioned in the rule, on both the left-hand side (LHS) and right-hand side (RHS). For example, if we want to find rules that apply to at least 7,500 transactions, then we would set the support to 7,500/75,000, or 0.1. The support can be calculated from the support count, which is the frequency of occurrence of an itemset. It is the number of times that the itemset is found in a purchase/transaction.

**Confidence:** The confidence is essentially an indication of how often a rule is true and is a fraction of transactions for which the rule is true. It is the fraction/percentage of the itemset on the RHS in transactions that contain the itemset on the LHS. In other words, it is the number of transactions for which this rule is true divided by the number of transactions on the LHS that satisfy the "if" condition. It indicates how confident we can be that the RHS itemset will be in a transaction if the LHS itemset is in a transaction.

**Lift:** The lift ratio helps us determine if items are being bought at random or not. It tells us the factor by which the probability of finding an item A in the basket has increased (or decreased) once we know that an item B is in the basket. The higher the lift, the greater the chances of the RHS itemset being in the basket given that the LHS itemset is in the basket. A lift of 1 tells us that the probability of occurrence for the two events are independent of each other and the appearance of one item in the basket does not affect whether another certain item will be in the basket. A lift greater than 1 indicates a positive dependence and a complementary effect on the items, while a lift less than 1 indiciates a negative dependence and a substitution effect on the items.

```{r}
# Let's start out by keeping things simple and make rules containing only 2-3 items.
basket_rules <- apriori(trans_data, parameter = list(sup = 0.04, conf = 0.5, target="rules", minlen = 2, maxlen = 3))
# Removing rendunant rules (rules that are a subset of another)
basket_rules <- basket_rules[!is.redundant(basket_rules)]
# Checking number of rules
length(basket_rules)
# Looking at all 16 rules
inspect(basket_rules)

# Now I will sort the rules from highest lift to lowest lift, and then take a look at the top 7 rules.
basket_rules_sorted <- sort(basket_rules, by = "lift")
inspect(basket_rules_sorted[1:7])
```

From these rules, we see that the top 3 involve the opera cake, apricot danish, and the cherry tart. These three items are found together in about 4.11% of baskets  The first rules tells us that shoppers who purchase an opera cake and an apricot danish are about 10.3 times more likely to purchase a cherry tart than the average shopper. Rules 4 and 5 indicate that chocolate coffee and chocolate cake are found together in about 4.4% of baskets. These shoppers must really like chocolate! Shoppers who purchase chocolate coffee are about 6.3 times more likely to purchase a chocolate cake.

Below you will find a simple visualization for these top 7 rules. The darker the red color, the higher the lift and the larger the size of the circle, the higher the support with the grey arrows connecting the circles to the item name. We could use this plot with even more rules for a better comparison, however one needs to be careful because the more rules included, the messier the plot could become and therefore the more difficult it could be to read and interpret.

```{r}
#install.packages("arulesViz")
library(arulesViz)
set.seed(200)
plot(basket_rules_sorted[1:7], method = "graph")
```

Finally, we will inspect rules with just 4 items included (three items in the 'if' statement and one item in the 'then' statement).

```{r}
basket_rules4 <- apriori(trans_data, parameter = list(sup = 0.02, conf = 0.5, target="rules", minlen = 4, maxlen = 4))
# Removing rendunant rules (rules that are a subset of another)
basket_rules4 <- basket_rules4[!is.redundant(basket_rules4)]
# Checking number of rules
length(basket_rules4)
# Looking at all 16 rules
inspect(basket_rules4)

# Now I will sort the rules from highest lift to lowest lift, and then take a look at the top 7 rules.
basket_rules4_sorted <- sort(basket_rules4, by = "lift")
inspect(basket_rules4_sorted[1:7])
```

We see that the first rule is quite different than the other 6. The first rule says that shoppers who purchase an apple tart, apple danish, and cherry soda are about 14.8 times more likely to purchase an apple croissant than the average shopper. The other remainng rules just about all involve something lemon-flavored and raspberry-flavored lemonades and cookies. It makes sense that a customer who is buying something involving those two flavors has a better chance of purchasing something else involving those flavors, like raspberry lemonade or a raspberry cookie.

I hope you have enjoyed this analysis! Thanks for reading!
